Relationship between solar wind, $D_{st}$, and plasmasphere mass density on one-hour time scales

### 2015-09-01 paper.tex
Paper updates; code updates

\documentclass[10pt,twocolumn]{article}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{amssymb,amsmath}
\usepackage{nicefrac}
\usepackage[usenames,dvipsnames]{color}
\usepackage[colorlinks,citecolor=RedViolet,urlcolor=blue]{hyperref}
\usepackage{doi}
\usepackage{setspace}
\usepackage[paperwidth=8.5in, paperheight=11in,top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage[compact]{titlesec}
\usepackage{abstract}
\usepackage[position=t,singlelinecheck=off,justification=raggedleft]{subcaption}

\usepackage[authoryear,square]{natbib}
\pagestyle{fancy}
\fancyhead[R]{Veibell \thepage}
\fancyhead[L]{}

@@

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period and 1991 and found that the average ion mass density, $M=\rho/n_e$, depended on geomagnetic activity, with lower $D_{st}$ values corresponding to slightly average ion mass densities.  They also found a $K_p$ dependence between the average ion mass and 1.5- and 3-day averages of $K_p$ were compared to $M$.  Most measurements occurred in the MLT range of 12:00 and 18:00 MLT in the plasma trough (in equatorial plane?)

\cite{Takahashi2010} developed a mass density dataset using measurements from the GOES satellites over the years 1980 through 1992, with most measurements in the range $L=6.8\pm0.2$ The equatorial mass density, $\rho_eq$ was estimated from the estimated mass density a power law dependence on the geocentric radius $R$, $\rho=\rho_{eq}(LR_E/R)$. A key result was that when daily-averaged measurements were considered, an enhancment in $\rho_{eq}$ appeared, indicating the possibility of a geomagnetic storm control over $\rho_{eq}$.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset associated with \cite{Denton}, with data available from 1980-1991; all other parameters are from \cite{Reconstruction} over the time range of 1972-2013, which are on a 1-hour time grid. $\rho_{eq}$ is the inferred mass density based on the 3rd harmonic frequency of magnetic field measurements.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hourly range was used.  Fill values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple Geostationary Operational Environmental Satellites (GOES) satellites at the same time in the \cite{Denton} dataset, the value from GOES 7 was selected.  In the dataset, measurements from GOES 7 had the longest time span of coverage.  An overview of the data is given in Figure~\ref{AllData}.

@@

In this work events defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated in by horizontal dashed lines in the respective panels of Figure~\ref{AllData}.  In finding events, all fill values were replaced with linearly interpolated values.  Figure \ref{AllData} shows the values of $B_z$, $V_{SW}$, $D_{st}$, $F_{10.7}$, and $\rho{eq}$ for the duration of the dataset.

@@

Figure~\ref{ccplot:a} shows the long-term trends of $log(\rho_{eq})$ and $F_{10.7}$ computed using the median values in 27-day (non-overalpping?) windows using all non--fill hourly values.  A scatter plot of these two lines is shown in Figure~\ref{ccplot:b}.  The linear correlation is found to be $0.87$.  In comparison, \cite{Takahashi2010} found a correlation coefficient of $0.94$ using measurements from all satellites in the time interval of 1980 through 1991.

@@

Two types of events are considered. The first is a drop in $D_{st}$ below the threshold of $-40nT$.  The second is an increase in $\rho_{eq}$ above 40~$amu/cm^3$.

Our first analysis uses data only from the time interval 1989-1991.  In this interval, there were (?) $D_{st}$ events and (?) $\rho_{eq}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians accross events.)  Error bars for each parameter were computed using (2x ?) the standard deviation of the number of values used in computing the median dividied by the square root of the number of values used.

In the upper panel of \label{DailyAverages}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48~nT$.  Consistent with \cite{Takahashi2010}, $D_{st}$ events corrspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is on the order of $~2 amu/cm^3$ instead of $~10 amu/cm^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for $D_{st}$, the number of measurements used in computing the medians is much larger as we use all available $D_{st}$ measurements instead of restricting to only values where a non-fill $\rho_{eq}$ value existed.

@@

This method finds 669 such periods between May 1983 and August 1991 with an average duration of 9 hours and a median duration of 3 hours. Figure \ref{Storms} shows the average values of all events over a window of 24 hours before onset and 48 hours after. Figure \ref{Storms}a shows events selected by looking for event onsets where $D_{st}$ crossed a threshold of $-40nT$. The dashed red lines indicate plus and minus one standard deviation of values from all events that went into the average. The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack. A subset of longer duration events will be looked at later.

This figure shows a definite spike in the $Z$ component of the magnetic field, as well as the defined drop in $DST$, but no obvious change in mass density at an hourly timescale. This points to an issue with only looking at long-timescale trends between density and $D_{st}$, and allows for the possibility that other factors are influencing the long term correlation since there's no obvious connection on a short timescale. One possibility is that, as suggested in \cite{Takahashi2010}, $F_{10.7}$ plays a significant role in driving long term density values which biases the long term correlation of density and $D_{st}$.

\subsection{Mass Density Events}
Figure \ref{Storms}b shows this same algorithm, but looking for a rise in mass density over a value of 40$g/cm^3$. This results in 130 events with a mean duration of 32 hours and a median duration of 17 hours, marked in red on the left figure.

@@

Neither of these seem to indicate anything too significant, so looking at $D_{st}$ events instead to look for something that causes a significant change in Mass Density results in Figure \ref{Dspec}. This shows that by either looking only at $D_{st}$ events that last longer than an hour (left) or at events where the onset condition is $D_{st}<-80nT$, a spike in mass density is seen, but also a definite lack of data availability to the point where that spike may be coming from less than five of the total 143 events.

Unfortunately there are no events in this time frame that are longer than 12 hours with $D_{st}$ minima lower than $-80nT$ that have existing mass density data around onset, so an analysis of this particular relationship can't be made.

@@

\section{F10.7 dependence}
In an effort to analyze the dependence of $\rho_{eq}$ on $F_{10.7}$, a few tests were performed. \cite{Takahashi2010} mention a strong correlation between the two. The long term correlation could be a bias for $D_{st}$'s effects, so a linear model was created, recreating mass density purely from $F_{10.7}$ in the form of $\rho_{eq}(t)=A*F_{10.7}(t)$. This re-created $\rho_{eq}$ shows around a 45\% correlation with the actual $\rho_{eq}$, suggesting a strong influence, while doing the same procedure with $DST$ shows only a 20-25\% correlation.

@@

\section{Appendix}
\subsection{Bias}
While attempting to reproduce Figure 11 from \cite{Takahashi2010}, an ambiguity in data-handling was found. It is unknown how they got from hourly $\rho_{eq}$ to daily medians, whether in one step or two steps, and whether the hourly medians are an hour ahead of each hour grid point, or a median of points a half hour to either side, so attempts were made to reproduce all possibilities and compare in order to find a data-handling method with the least bias. These attempts are shown in Figure \ref{blockmedian}. This also shows the effect of only considering events with a minimum before noon. All show a similar spike in $\rho_{eq}$ at the time of $D_{st}$ minimum, though none quite reproduce the exact values seen in \cite{Takahashi2010}. Figure \ref{allstorms} shows where the median falls for all of the found events during that period, as a sanity check for why the prior work would get median values nearing $30 amu/cm^3$. Both authors of this paper conducted independent trials to confirm that the data and analysis don't quite yield the results of Figure 11 in \cite{Takahashi2010} with the processes described in that paper, but the same general trend is still seen.

@@

\newpage
\footnotesize
\bibliographystyle{vplainnat}

\bibliography{reportbib}

@@

\begin{figure}[htp]
\centering
\includegraphics[scale=0.45]{paperfigures/alldata.eps}
\caption{Overview of data used in this study. The top four panels show parameters from \cite{Reconstruction} and the bottom panel is based on $\rho_{eq}$ from \cite{Denton} after interpolation and averaging described in the text. Dashed horizontal lines in the $D_{st}$ and $\rho_{eq}$ panels indicate sample event cutoff thresholds of $D_{st}=-50~nT$ and $\rho_{eq}=40~amu/cm^3$}.
\label{AllData}
\end{figure}

\begin{figure}[htp!]
\centering
\subcaptionbox{\label{ccplot:a}}
{\includegraphics[scale=0.42]{paperfigures/F107MDAllData.eps}}
\subcaptionbox{\label{ccplot:b}}
{\includegraphics[scale=0.42]{paperfigures/ccplot.eps}}
\caption{(a) All data of $F_{10.7\_27d}$ and $\rho_{eq\_27d}$. (b) Correlation between $log(\rho_{eq})$ and $F_{10.7}$ at hour, day, and 27-day intervals; compare to \cite{Takahashi2010} Fig. 14}
\label{ccplot}
\end{figure}
\clearpage

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-dst.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-mass.eps}
\caption{(a) Average of solar wind and near-Earth measurements around the time $D_{st}$ crossed below $-50~nT$ onset. (b) Same as (a) except around time intervals where mass density crossed above $40~amu/cm^3$.  The bottom panel shows the number of mass density data points that were used for the average.}
\label{Storms}
\end{figure}

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-dd12.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-md12.eps}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ stayed below $-50~nT$ for at least 12~hours. (b) Same as Figure \ref{Storms}(b) except for constrain that $\rho_{eq}$ stayed above $50~amu/cm^3$ for at least 12 hours.}
\label{Mspec}
\end{figure}
\clearpage

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-d80.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-m70.eps}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ crossed below $-80nT$ (b) Same as Figure \ref{Storms}(b) except for constrain that $\rho_{eq}$ crossed above $70~amu/cm^3$.}
\label{Dspec}
\end{figure}

@@

\begin{figure}[htp]
\centering
\includegraphics[scale=0.5]{paperfigures/nansbyhour.eps}
\includegraphics[scale=0.5]{paperfigures/nansbyhour_storm.eps}
\caption{(a) Number of NaN points per hour of observation in the total data set (b) $\rho_{eq}$ data availability relative to average event hour in events where $D_{st}<-80nT$}
\label{nanperhour}
\end{figure}
\clearpage

>>>>

\documentclass[10pt,twocolumn]{article}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{amssymb,amsmath}
\usepackage[usenames,dvipsnames]{color}
\usepackage[colorlinks,citecolor=RedViolet,urlcolor=blue]{hyperref}
\usepackage{doi}
\usepackage{setspace}
\usepackage[paperwidth=8.5in, paperheight=11in,top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage[compact]{titlesec}
\usepackage{abstract}
\usepackage[position=t,singlelinecheck=off,justification=raggedleft]{subcaption}
\usepackage[authoryear,square]{natbib}
\pagestyle{fancy}
\fancyhead[R]{Veibell \thepage}
\fancyhead[L]{}

@@

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period and 1991 and found that the average ion mass density, $M=\rho/n_e$, correlated with geomagnetic activity, with lower $D_{st}$ values corresponding to slightly average ion mass densities.  They also found a relationship between the average ion mass and 1.5- and 3-day averages of $K_p$ were compared to $M$.  Most measurements occurred in the MLT range of 12:00 and 18:00 in the plasma trough (in equatorial plane?)

\cite{Takahashi2010} developed a mass density dataset using measurements from the GOES satellites over the years 1980 through 1992, with most measurements in the range $L=6.8\pm0.2$ The equatorial mass density, $\rho_{eq}$ was estimated from the estimated mass density a power law dependence on the geocentric radius $R$, $\rho=\rho_{eq}(LR_E/R)$. A key result was that when daily-averaged measurements were considered, an enhancment in $\rho_{eq}$ appeared, indicating the possibility of a solar wind control over $\rho_{eq}$.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset associated with \cite{Denton}, with data available from 1980-1991; all other parameters are from \cite{Reconstruction} over the time range of 1972-2013, which are on a 1-hour time grid. $\rho_{eq}$ is the inferred mass density based on the 3rd harmonic frequency of magnetic field measurements.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hourly range was used.  Fill values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple Geostationary Operational Environmental Satellites (GOES) satellites at the same time in the \cite{Denton} dataset, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data is given in Figure~\ref{AllData}.

@@

In this work, events defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated in by horizontal dashed lines in the respective panels of Figure~\ref{AllData}.  In finding events, all fill values were replaced with linearly interpolated values.  Figure \ref{AllData} shows the values of $B_z$, $V_{sw}$, $D_{st}$, $F_{10.7}$, and $\rho_{eq}$ for the duration of the dataset.

@@

Figure~\ref{ccplot:a} shows the long-term trends of $log(\rho_{eq})$ and $F_{10.7}$ computed using the median values in 27-day (non-overalpping?) windows using all non--fill hourly values.  A scatter plot of these two lines is shown in Figure~\ref{ccplot:b}.  The linear correlation is found to be $0.87$ (? plot says 0.94!).  In comparison, \cite{Takahashi2010} found a correlation coefficient of $0.94$ using measurements from all satellites in the time interval of 1980 through 1991.

@@

Two types of events are considered. The first is a drop in $D_{st}$ below the threshold of $-40$~nT.  The second is an increase in $\rho_{eq}$ above $40$~amu/cm$^3$.

Our first analysis uses data only from the time interval 1989-1991.  In this interval, there were (?) $D_{st}$ events and (?) $\rho_{eq}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians accross events.)  Error bars for each parameter were computed using the standard deviation (2x ?) of values used in computing the median dividied by the square root of the number of values used.

In the upper panel of \label{DailyAverages}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events corrspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is on the order of $\sim2$~amu/cm$^3$ instead of $\sim10$~amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for $D_{st}$, the number of measurements used in computing the medians is much larger as we use all available $D_{st}$ measurements instead of restricting to only values where a non-fill $\rho_{eq}$ value existed.

@@

A total of 669 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours. Figure \ref{Storms} shows the average values of all events over a window of 24 hours before onset and 48 hours after. Figure \ref{Storms}a shows events selected by looking for event onsets where $D_{st}$ crossed a threshold of $-40$~nT. The dashed red lines indicate plus and minus one standard deviation of values from all events that went into the average. The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack. A subset of longer duration events will be looked at later.

This figure shows a definite spike in the $Z$ component of the magnetic field, as well as the defined drop in $D_{st}$, but no obvious change in mass density at an hourly timescale. This points to an issue with only looking at long-timescale trends between density and $D_{st}$, and allows for the possibility that other factors are influencing the long term correlation since there's no obvious connection on a short timescale. One possibility is that, as suggested in \cite{Takahashi2010}, $F_{10.7}$ plays a significant role in driving long term density values which biases the long term correlation of density and $D_{st}$.

\subsection{Mass Density Events}
Figure \ref{Storms}b shows this same algorithm, but looking for a rise in mass density over a value of $40$~amu/cm$^3$. This results in 130 events with a mean duration of 32 hours and a median duration of 17 hours, marked in red on the left figure.

@@

Neither of these seem to indicate anything too significant, so looking at $D_{st}$ events instead to look for something that causes a significant change in Mass Density results in Figure \ref{Dspec}. This shows that by either looking only at $D_{st}$ events that last longer than an hour (left) or at events where the onset condition is $D_{st}<-80$~nT, a spike in mass density is seen, but also a definite lack of data availability to the point where that spike may be coming from less than five of the total 143 events.

Unfortunately there are no events in this time frame that are longer than 12 hours with $D_{st}$ minima lower than $-80$~nT that have existing mass density data around onset, so an analysis of this particular relationship can't be made.

@@

\section{F10.7 dependence}
In an effort to analyze the dependence of $\rho_{eq}$ on $F_{10.7}$, a few tests were performed. \cite{Takahashi2010} mention a strong correlation between the two. The long term correlation could be a bias for $D_{st}$'s effects, so a linear model was created, recreating mass density purely from $F_{10.7}$ in the form of $\rho_{eq}(t)=A\cdot F_{10.7}(t)$. This re-created $\rho_{eq}$ shows around a 45\% correlation with the actual $\rho_{eq}$, suggesting a strong influence, while doing the same procedure with $D_{st}$ shows only a 20-25\% correlation.

@@

\section{Appendix}
\subsection{Bias}
While attempting to reproduce Figure 11 from \cite{Takahashi2010}, an ambiguity in data-handling was found. It is unknown how they got from hourly $\rho_{eq}$ to daily medians, whether in one step or two steps, and whether the hourly medians are an hour ahead of each hour grid point, or a median of points a half hour to either side, so attempts were made to reproduce all possibilities and compare in order to find a data-handling method with the least bias. These attempts are shown in Figure \ref{blockmedian}. This also shows the effect of only considering events with a minimum before noon. All show a similar spike in $\rho_{eq}$ at the time of $D_{st}$ minimum, though none quite reproduce the exact values seen in \cite{Takahashi2010}. Figure \ref{allstorms} shows where the median falls for all of the found events during that period, as a sanity check for why the prior work would get median values nearing $30$~amu/cm$^3$. Both authors of this paper conducted independent trials to confirm that the data and analysis don't quite yield the results of Figure 11 in \cite{Takahashi2010} with the processes described in that paper, but the same general trend is still seen.

@@

\newpage
\footnotesize
\bibliographystyle{plainnat}
\bibliography{reportbib}

@@

\begin{figure}[htp]
\centering
\includegraphics[scale=0.45]{paperfigures/alldata.eps}
\caption{Overview of data used in this study. The top four panels show parameters from \cite{Reconstruction} and the bottom panel is based on $\rho_{eq}$ from \cite{Denton} after interpolation and averaging described in the text. Dashed horizontal lines in the $D_{st}$ and $\rho_{eq}$ panels indicate sample event cutoff thresholds of $D_{st}=-50$~nT and $\rho_{eq}=40$~amu/cm$^3$}.
\label{AllData}
\end{figure}

\begin{figure}[htp!]
\centering
\subcaptionbox{\label{ccplot:a}}
{\includegraphics[scale=0.42]{paperfigures/F107MDAllData.eps}}
\subcaptionbox{\label{ccplot:b}}
{\includegraphics[scale=0.42]{paperfigures/ccplot.eps}}
\caption{(a) 27-day averages of $F_{10.7}$ and $\rho_{eq}$. (b) Correlation between $log(\rho_{eq})$ and $F_{10.7}$ using hour, day, and 27-day averaging intervals; compare to \cite{Takahashi2010} Fig.~14.}
\label{ccplot}
\end{figure}
\clearpage

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-dst.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-mass.eps}
\caption{(a) Average of solar wind and near-Earth measurements around the time $D_{st}$ crossed below $-50$~nT onset. (b) Same as (a) except around time intervals where mass density crossed above $40$~amu/cm$^3$.}
\label{Storms}
\end{figure}

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-dd12.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-md12.eps}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours. (b) Same as Figure \ref{Storms}(b) except for constrain that $\rho_{eq}$ stayed above $50$~amu/cm$^3$ for at least 12 hours.}
\label{Mspec}
\end{figure}
\clearpage

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-d80.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-m70.eps}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ crossed below $-80$~nT (b) Same as Figure \ref{Storms}(b) except for constrain that $\rho_{eq}$ crossed above $70~amu/cm^3$.}
\label{Dspec}
\end{figure}

@@

\begin{figure}[htp]
\centering
\includegraphics[scale=0.5]{paperfigures/nansbyhour.eps}
\includegraphics[scale=0.5]{paperfigures/nansbyhour_storm.eps}
\caption{(a) Number of NaN points per hour of observation in the total data set (b) $\rho_{eq}$ data availability relative to average event hour in events where $D_{st}<-80$~nT.}
\label{nanperhour}
\end{figure}
\clearpage
